Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-02-04 analysis/paper/paper.Rmd
note on intro
work on results

2.  Habitat fragmentation is a major source of environmental stress. **[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]**  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. Of particualar concern are the effects of fragmentation on species-rich tropical ecosystems. In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest >1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015]. Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H. acuminata*---are usually found at lower densities and in the more shaded forest understory. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants. We also surveyed plots regularly during the rainy season to identify any that flowered after the survey. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. The estimates of potential evapotranspiration were calculated using the Pennman-Monteith method, and then along with monthly precipitation were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. Positive SPEI values indicate wetter than average months<!-- instead: "...indicate the specified time period is wetter than average." because time frame doesn't have to be a month, yes?  -->, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (< -2) [@mckeeRelationshipDroughtFrequency1993]. Because *H. acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales. We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).

@@

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, referred to as the crossbasis function. This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This particular maximum lag time in respont to drought was chosen because *H. acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002]. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017]. In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.

@@

$$
g[\textrm{E}(y_{i})] = \alpha_j + f_1(x_i) + f_2(d_{i,t}, ..., d_{i,t-24}) \\
\alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
$$

Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details). Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex. A random effect of plot ID on the intercept is represented by $a_j$. For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$). A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure. The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences). When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.

------------------------------------------------------------------------

NOTE FOR EMILIO: In Gasparrini et al. 2017, the crossbasis function is defined like so:

$$
s(x_{i,t-l_0}, ...,x_{i,t-L}) = \sum_{l = l_0}^L f \cdot w(x_{i,t-l},l) 
$$

Where "the bi-dimensional dose-lag-response function $f \cdot w (x,l)$ is composed of two marginal functions: the standard dose-response function $f(x)$, and the additional lag-response function $w(l)$ that models the lag structure...". I like this summation notation because it clarifies that the function is two-dimensional, but I'm not sure it's mathematically appropriate to drop into the model equation.\
The crossbasis function could also be represented by something a lot more abstract like $f_2(X_i, L)$ where $X_i$ is the SPEI history and $L$ is a vector (0, ..., 36) describing the lag structure of $X_i$. This is closer to how the model is written in R. Let's talk about the clearest way to write and describe these models.

------------------------------------------------------------------------

Because DLNMs do not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

Statistical significance of crossbasis functions and covariates was determined by .... <!--# not sure yet! -->

@@

## DLNMs

>>>>

2.  Habitat fragmentation is a major source of environmental stress. **[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]** <!--# fragmentation results in increaed mortality due to increased stress. fragmentation itself is not the stressor -->Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. Of particular concern are the effects of fragmentation on species-rich tropical ecosystems. In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015]. Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H. acuminata*---are usually found at lower densities and in the more shaded forest understory. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** <!--# check data --> flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants. We also surveyed plots regularly during the rainy season to identify any that flowered after the survey. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. Plants that were not found for three consecutive surveys were considered dead. In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. The estimates of potential evapotranspiration were calculated using the Pennman-Monteith method, and then along with monthly precipitation were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993]. Because *H. acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales. We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).

@@

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function. This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. We chose 36 months as a maximum lag because *H. acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002]. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].

@@

$$
g[\textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
\alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
$$

Where $s_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details). The crossbasis function, $s_2(\cdot)$ can also be written:

$$
s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
$$

Where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$. These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodsimonn.ChapterSmoothers] and Gasparrini et al. [-@gasparriniPenalizedFrameworkDistributed2017] for more detail). Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex. A random effect of plot ID on the intercept is represented by $a_j$. For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$). A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure. The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences). When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity <!--# refrence Wood 2006? -->. We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].

Because the `dlnm` package does not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.

@@

Start with descriptive results of drought events---how many droughts, how severe, timing of droughts, etc.

## Survival

Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha. Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004. Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, will get exact numbers from data). The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.

### Continuous Forest

size: p \< 0.001 (figure = surv_cf_size.png)

plot: p \< 0.001

SPEI: p \< 0.001 (figure = surv_cf_spei.png)

Lags 6-0 (Aug--Feb) and 20-15 (Jun--Nov) show reduced survival with drought. Lags 14-7 (Dec - Jul) and 26-21 (Dec-May) show increased survival with drought. Lag 36-33 (Feb - May) shows quadratic relationship with increased survival in dry and wet conditions (this seems like it's gotta be spurious).

### 1-ha fragment

size: p \<0.001 (figure = surv_1ha_size.png)

plot: p = 0.258

SPEI: p \< 0.001 (figure = surv_1ha_spei.png)

Crossbasis is very flat with essentially no effect of SPEI beyond 6 months lag. Recent SPEI has bimodal effect with both low and high values reducing survival.

**Question**: Currently the figures are expected values for the average plant across both plots. This makes it look like survival is a lot higher in 1-ha fragments when in reality survival is exactly the same (0.95) because plants are smaller in 1-ha fragments. Should these plots instead be for an average sized plant within the habitat type?

## Growth

Plants were on average larger in continuous forest compared to 1-ha fragments. In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150. In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.

The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the dataset with mean sizes becoming more similar over time.

**NOTE**: I considered adding a covariate of year to account for yearly trends in growth, but since there is also a trend in SPEI over the time of this experiment, they were concurve (wiggly collinearity), and shouldn't both be included in model.

### Continuous Forest

size: p \< 0.001

plot: p \< 0.001

SPEI: p \< 0.001

Predicted results show only lagged effects of SPEI on growth (f(x) at lags 0 and 1 is flat). Wet conditions in preceding winter months improve growth (lags 4-2, Oct - Dec). Positive effect of drought on growth at lags 11-8 (Mar - Jun) and 34-24 (nearly the entirety of year t-2).

### 1-ha fragment

size: p \< 0.001

plot: p \< 0.001

SPEI: p \< 0.001

Crossbasis looks very similar to continuous forest but more extreme. SPEI has a stronger effect on growth in fragments. Wet conditions at lag 0 predict the greatest shrinkage of plants in fragments (∆size = -75 at SPEI \>2), but at lags 5-2 (Sept-Dec) high SPEI has a less negative effect on plant size. We only expect plant growth (∆size \> 0) in years when SPEI was low at lags 34-32

The size of plants significantly predicted their size in the following year in continuous forest (edf = 7.8, P \< 0.001 <!--# What stats to include?  reference df and F also?  Is this all better as a table? -->) and forest fragments (edf = 7.4, P \< 0.001). In continuous forest, plants under a size of 259 were predicted to grow from year t-1 to year t, and the fit line was below 1:1 for larger plants, indicating that larger plants tended to shrink, on average. In 1 ha fragments, the fit line followed a 1:1 line more closely, indicating that plants tended to stay the same size on average from year to year over the course of this study. <!--# something weird about backtransformation causing these strange results? --> Drought history also had a significant effect on plant size in both continuous forest (edf = 14.7, p \< 0.001) and fragments (edf = 17.8, p \< 0.001).

## Flowering

Probability of flowering increased with size and was generally higher in continuous forest compared to 1-ha fragments. The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest. The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.

### Continuous forest

size: p \< 0.001

plot: p\<0.001

SPEI: p \< 0.001

Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov). Drought at lags 13-7 reduced flowering probability
